---
title: 'IAVS-2026: "The Burning Question"'
author: "Gabriela Dezotti"
date: "2024-06"
lang: "en"
format: 
  html:
   theme: cosmo
   number-sections: true
   toc: true
   code-fold: true
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
  cache: true
---

# Objective
This document presents the analyses developed for my poster at the 68th IAVS Symposium. Using data from a long-term experiment in a Cerrado area, I examine how forb community richness and cover change over time in response to fire suppression. The workflow includes data organization, analysis, and figure generation for the poster.

# Methods
Forb classification follows the document [20260420_GrowthFormsTombador_AFidelis](https://docs.google.com/spreadsheets/d/1QtTrJXviJqreZNYnwPIZKfGR0B7WD2L9/edit?usp=sharing&ouid=105834115709745946255&rtpof=true&sd=true),which lists species and their growth forms for the RNST area. I compare the fire-excluded and mid-dry-season treatments using only rainy-season surveys to reduce the confounding effect of seasonality on community structure. Because the rainy season is the period of highest productivity in the Cerrado, this approach also increases the likelihood of detecting species that are only visible during this period.

# Data Preparation

I first load the packages required for data handling, modelling, and plotting. These include `tidyverse`, `glmmTMB`, `DHARMa`, `ggeffects`, `extrafont`, and `emmeans`.
Next, I recode treatment labels to make figures more interpretable and define the colour palette used throughout the poster.
I then read the raw dataset, which contains information on forb cover and species richness through 2016 to 2025.

```{r}
#| label: setup
#| include: false
library(here)
library(tidyverse)
library(glmmTMB)
library(DHARMa)
library(ggeffects)
library(extrafont)
library(emmeans)

  recode_treatment = function(x) {
  recode(x,
    "BF" = "Frequent fire",
    "FE" = "Fire suppression"
  )
}

cols = c(
  "Frequent fire" = "#a6531b",
  "Fire suppression" = "#0e0d0d" 
)

fills = c(
  "Frequent fire" = "#e65c00",
  "Fire suppression" = "#555555"
)

loadfonts()

year_levels = c("2016", "2017", "2018", "2019", "2020", "2022", "2023", "2024", "2025")
```

```{r}
#| label: theme-poster
#| include: false
#| 
poster = function(show_legend = FALSE, legend_pos = c(0.48, 0.93)) {
  theme_bw(base_family = "Aptos") +
    theme(
      axis.title = element_text(face = "bold", size = 20),
      axis.text.x = element_text(size = 16, color = "black"),
      axis.text.y = element_text(size = 16, color = "black"),
      legend.text = element_text(size = 16),
      legend.position = if (show_legend) legend_pos else "none",
      legend.justification = c(0.5, 1),
      legend.direction = "horizontal",
      plot.margin = margin(t = 15, r = 15, b = 15, l = 15),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(color = "#4a3b32", fill = NA, linewidth = 1.5),
      plot.background = element_rect(fill = "transparent", color = NA),
      panel.background = element_rect(fill = "transparent", color = NA),
      legend.background = element_rect(fill = "transparent", color = NA)
    )
}
```

```{r}
#| label: load-data
#| include: true

data = read.table(
  here("data", "data_forbs_iavs_2026.txt"), 
                   header = TRUE, 
                   stringsAsFactors = FALSE, 
                   sep="\t") 

glimpse(data)    
```

# Data Wrangling
Here, I summarise the species-level data by date, treatment, plot, and observer. For each sampling unit, I calculate total forb cover and species richness, then extract year from the survey date and store it as a factor.
```{r}
#| label: data-prep
#| include: true

  cover = data |>

  group_by(date, treatment, plots, observer) |>
  summarise(
    totalCover = sum(cover, na.rm = TRUE),
    richness = n_distinct(species),
    .groups = "drop"
  ) |>

  mutate(
    year = as.numeric(substr(as.character(date), 1, 4))
  ) |>
  
  select(-date) |>
  mutate(
    year = factor(year)
  )

glimpse(cover)
```

## Cover
To test how total forb cover changes over time under different fire regimes, I fit a GLMM with treatment, year, and their interaction as fixed effects.
I used a Tweedie distribution with a log link to accommodate the distribution of cover values, and included observer and plot as random intercepts to account for repeated measurements and observer-related variation

### Model fitting
```{r}
#| label: fit-cover-model

glmm_cover = glmmTMB(
  totalCover ~ treatment * year + (1 | observer) + (1 | plots),
   family = tweedie(link = "log"),
   data = cover
)

summary(glmm_cover)
```

Total forb cover was lower under fire suppression in the reference year (2016). The treatment-by-year interaction indicates that differences between fire regimes were not constant through time.

### Model diagnostics
I assessed model fit using simulated residuals from DHARMa, including graphical inspection and tests for dispersion and residual uniformity. These checks help evaluate whether the model adequately captures the distributional structure of the data.

```{r}
#| label: diag-cover-model

res_cover <- simulateResiduals(glmm_cover)
plot(res_cover)
testDispersion(res_cover)
testUniformity(res_cover) 
```

```{r}
#| label: trends-cover-model

emm_cover =
 emmeans(glmm_cover, ~ treatment | year,
 type = "response")

contrast_cover <- pairs(emm_cover, adjust = "holm")

summary(contrast_cover, infer = TRUE)
```

Forb cover was already lower under fire suppression in the first sampled year and remained lower in most subsequent years.
The magnitude of this difference varied over time, but the pattern indicates an early decline in cover following fire exclusion.

## Richness
To test how forb species richness changes over time under different fire regimes, I fitted a generalized linear mixed model with treatment, year, and their interaction as fixed effects. I included plot as a random intercept to account for repeated measurements within plots. I used a Conway–Maxwell–Poisson distribution with a log link to model richness.

### Model fitting
```{r}
#| label: fit-richness-model
#| include: true
  glmm_rich = glmmTMB(
    richness ~ treatment * year + (1 | plots),
    family = compois(link = "log"),
    data = cover
  )

  summary(glmm_rich)
```

Forb richness was lower under fire suppression in the baseline year. The interaction between treatment and year indicates that differences between fire regimes were not constant through time, so I used estimated marginal means to identify in which years the two treatments differed.

### Model diagnostics
```{r}
#| label: diag-richness-model
#| include: true

 res_rich = simulateResiduals(glmm_rich)
 plot(res_rich)
 testDispersion(res_rich)
 testUniformity(res_rich) 

```

To evaluate when richness differed between fire regimes, I estimated marginal means for each treatment within each year and performed pairwise contrasts with Holm adjustment. This step directly tests when fire-suppressed communities begin to diverge from frequently burned communities.

```{r}
#| label: pairs-richness-model

emm_rich = 
emmeans(glmm_rich, ~ treatment | year, type = "response")

rich_tbl =
 pairs(emm_rich, adjust = "holm")

  summary(rich_tbl,infer = TRUE)
```

Forb richness was higher under frequent fire in several years, but the difference was not consistent across all surveys. 
**Richness differences became clearer and more recurrent from 2020 onward.**

# Figures

## Data preparation for plotting
```{r}
#| label: build-plot-data

rich_sum =
 cover |>
  group_by(treatment, year) |>
  summarise(mean_rich = mean(richness, na.rm = TRUE), .groups = "drop") |>
  mutate(
    treatment = recode_treatment(treatment))

cover_sum =
 cover |>
  group_by(treatment, year) |>
  summarise(mean_cover = mean(totalCover, na.rm = TRUE), .groups = "drop") |>
  mutate(
    treatment = recode_treatment(treatment),
    year_plot = factor(as.character(year), levels = year_levels)
  )

pred_rich = ggpredict(
  glmm_rich,
  terms = c("year", "treatment"),
  bias_correction = TRUE
) |>
  as.data.frame() |>
  mutate(
    group = recode_treatment(group)
  )

pred_cover <- ggpredict(
  glmm_cover,
  terms = c("year", "treatment"),
  bias_correction = TRUE
) |>
  as.data.frame() |>
  mutate(
    group = recode_treatment(group)
  )

```


### Richness
```{r}
#| label: plot-richness
#| fig-width: 9
#| fig-height: 4.5

p_rich =
 ggplot() +
  geom_ribbon(
    data = pred_rich,
    aes(x = x, ymin = conf.low, ymax = conf.high, fill = group, group = group),
    alpha = 0.18,
    color = NA
  ) +

  geom_line(
    data = pred_rich,
    aes(x = x, y = predicted, color = group, group = group),
    linewidth = 1.8
  ) +

   scale_x_discrete(drop = FALSE, expand = expansion(mult = c(0, 0.02))) + 
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.15)),
  limits = c(0, NA)) +

  scale_color_manual(values = cols) +
  scale_fill_manual(values = fills) +
  labs(x = "Year", y = "Richness (spp/m²)", color = NULL) +
  poster(show_legend = FALSE) + 
  theme(axis.text.x = element_text(margin = margin(t = 5), size = 18))

  ggsave("richness.png", p_rich,
   width = 9, height = 4.5,
   dpi = 300, bg = "transparent")
```

Estimated richness trajectories indicate a slower and less consistent response than cover. Differences between fire regimes were small or unclear in the earlier surveys, but became more evident in the later years of the study.

### Cover
```{r}
#| label: plot-cover
#| fig-width: 9
#| fig-height: 4.5

p_cover <- ggplot() +
  geom_ribbon(
    data = pred_cover,
    aes(x = x, ymin = conf.low, ymax = conf.high, fill = group, group = group),
    alpha = 0.18,
    color = NA
  ) +
  geom_line(
    data = pred_cover,
    aes(x = x, y = predicted, color = group, group = group),
    linewidth = 1.8
  ) +
  scale_x_discrete(drop = FALSE, expand = expansion(mult = c(0, 0.02))) + 
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.15))) +
  
  scale_color_manual(values = cols) +
  scale_fill_manual(values = fills) +

  labs(x = "Year", y = "Cover (%)", color = NULL) +
  guides(
    fill = "none",
    color = guide_legend(
      override.aes = list(size = 4)
    )) +
  poster(show_legend = FALSE) +
  theme(axis.text.x = element_text(margin = margin(t = 12), size = 18))

  ggsave(
  filename = "cover.png", 
  plot = p_cover, 
  width = 9,      
  height = 4.5, 
  dpi = 300, 
  bg = "transparent" 
)

```
Estimated cover trajectories indicate an early and persistent effect of fire regime. Forb cover was generally higher under frequent fire than under fire suppression throughout the study, although the magnitude of this difference varied among years.