---
title: 'IAVS-2026: "The Burning Question"'
author: "Gabriela Dezotti"
date: "2024-05"
lang: "en"
format: html
  theme: cosmo
  number-sections: true
  toc: true
  code-fold: true
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
  cache: true
editor: visual
---

# Objective

This document presents the analyses developed for my poster at the 68th IAVS Symposium. Using data from a long-term experiment in a Cerrado area, I quantify how the forb community structure changes over time in response to fire suppression. The workflow includes data organization, exploratory analyses, and figure generation for inclusion in the poster.

For defining forbs I've followed the document [20260420_GrowthFormsTombador_AFidelis](https://docs.google.com/spreadsheets/d/1QtTrJXviJqreZNYnwPIZKfGR0B7WD2L9/edit?usp=sharing&ouid=105834115709745946255&rtpof=true&sd=true), which is a list of species and their growth forms for the RNST Area.

I will be comparing fire excluded and mid-dry treatment and will be only using surveys carried out in the rainy season to avoid the confounding effect of seasonality on the community structure. The rainy season is also the period of higher productivity in the Cerrado, which is when we expect to see some species that only appear in tis period (Dezotti et al., in prep).

# Data Preparation

First, I load the necessary packages (tidyverse, vegan, glmmTMB) and read the data from a text file. The dataset contains information on plant cover and species richness.

```{r}
#| label: setup
#| include: false

library(here)
library(tidyverse)
library(glmmTMB)
library(DHARMa)
library(ggeffects)
library(extrafont)
library(emmeans)

  recode_treatment = function(x) {
  recode(x,
    "BF" = "Frequent fire",
    "FE" = "Fire suppression"
  )
}

cols = c(
  "Frequent fire" = "#a6531b",
  "Fire suppression" = "#0e0d0d" 
)

fills = c(
  "Frequent fire" = "#e65c00",
  "Fire suppression" = "#555555"
)

loadfonts()

poster = function(show_legend = FALSE, legend_pos = c(0.48, 0.93)) {
  theme_bw(base_family = "Aptos") +
    theme(
      axis.title = element_text(face = "bold", size = 20),
      axis.text.x = element_text(size = 16, color = "black"),
      axis.text.y = element_text(size = 16, color = "black"),
      legend.text = element_text(size = 16),
      legend.position = if (show_legend) legend_pos else "none",
      legend.justification = c(0.5, 1),
      legend.direction = "horizontal",
      plot.margin = margin(t = 15, r = 15, b = 15, l = 15),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(color = "#4a3b32", fill = NA, linewidth = 1.3),
      plot.background = element_rect(fill = "transparent", color = NA),
      panel.background = element_rect(fill = "transparent", color = NA),
      legend.background = element_rect(fill = "transparent", color = NA)
    )
}

year_levels = c("2016", "2017", "2018", "2019", "2020", "2022", "2023", "2024", "2025")

```

```{r}
#| label: load-data
#| include: true

data = read.table(here("Savanna.txt"), 
                   header = T, 
                   stringsAsFactors = , 
                   sep="\t") 

glimpse(data)    
```

# Data Wrangling

```{r}
#| label: data-prep
#| include: true

  cover = data |>

  group_by(date, treatment, plots, observer) |>
  summarise(
    totalCover = sum(cover, na.rm = TRUE),
    richness = n_distinct(species),
    .groups = "drop"
  ) |>

  mutate(
    year = as.numeric(substr(as.character(date), 1, 4))
  ) |>
  
  select(-date) |>
  mutate(
    year = factor(year)
  )

glimpse(cover)
```

## Cover

```{r}
#| label: fit-cover-model

glmm_cover <-  glmmTMB(
 totalCover ~ treatment * year + (1 | observer) + (1 | plots),
 family = tweedie(link = "log"),
 data = cover
)

summary(glmm_cover)
```

```{r}
#| label: diag-cover-model

res_cover <- simulateResiduals(glmm_cover)
plot(res_cover)
testDispersion(res_cover)
testUniformity(res_cover) 
```

```{r}
#| label: trends-cover-model

emm_cover =
 emmeans(glmm_cover, ~ treatment | year,
 type = "response")

contrast_cover <- pairs(emm_cover, adjust = "holm")

summary(contrast_cover, infer = TRUE)
```

## Richness

```{r}
#| label: fit-richness-model
#| include: true
  glmm_rich = glmmTMB(
    richness ~ treatment * year + (1 | plots),
    family = compois(link = "log"),
    data = cover
  )

  summary(glmm_rich)
```

```{r}
#| label: diag-richness-model
#| include: true

 res_rich <- simulateResiduals(glmm_rich)
 plot(res_rich)
 testDispersion(res_rich)
 testUniformity(res_rich) 

```

```{r}
#| label: pairs-richness-model

emm_rich = 
emmeans(glmm_rich, ~ treatment | year, type = "response")

rich_tbl =
 pairs(emm_rich, adjust = "holm") |>
  summary(infer = TRUE)

```

# Figures

```{r}
#| label: build-plot-data

rich_sum =
 cover |>
  group_by(treatment, year) |>
  summarise(mean_rich = mean(richness, na.rm = TRUE), .groups = "drop") |>
  mutate(
    treatment = recode_treatment(treatment))

cover_sum =
 cover |>
  group_by(treatment, year) |>
  summarise(mean_cover = mean(totalCover, na.rm = TRUE), .groups = "drop") |>
  mutate(
    treatment = recode_treatment(treatment),
    year_plot = factor(as.character(year), levels = year_levels)
  )

pred_rich = ggpredict(
  glmm_rich,
  terms = c("year", "treatment"),
  bias_correction = TRUE
) |>
  as.data.frame() |>
  mutate(
    group = recode_treatment(group)
  )

pred_cover <- ggpredict(
  glmm_cover,
  terms = c("year", "treatment"),
  bias_correction = TRUE
) |>
  as.data.frame() |>
  mutate(
    group = recode_treatment(group)
  )

```

```{r}
#| label: plot-richness
#| fig-width: 9
#| fig-height: 4.5

p_rich =
 ggplot() +
  geom_ribbon(
    data = pred_rich,
    aes(x = x, ymin = conf.low, ymax = conf.high, fill = group, group = group),
    alpha = 0.18,
    color = NA
  ) +

  geom_line(
    data = pred_rich,
    aes(x = x, y = predicted, color = group, group = group),
    linewidth = 1.8
  ) +

   scale_x_discrete(drop = FALSE, expand = expansion(mult = c(0, 0.02))) + 
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.15)),
  limits = c(0, NA)) +

  scale_color_manual(values = cols) +
  scale_fill_manual(values = fills) +
  labs(x = "Year", y = "Richness (spp/m²)", color = NULL) +
  poster(show_legend = FALSE) + 
  theme(axis.text.x = element_text(margin = margin(t = 5), size = 18))

  ggsave("richness.png", p_rich,
   width = 9, height = 4.5,
   dpi = 300, bg = "transparent")
```

```{r}
#| label: plot-cover
#| fig-width: 9
#| fig-height: 4.5

p_cover <- ggplot() +
  geom_ribbon(
    data = pred_cover,
    aes(x = x, ymin = conf.low, ymax = conf.high, fill = group, group = group),
    alpha = 0.18,
    color = NA
  ) +
  geom_line(
    data = pred_cover,
    aes(x = x, y = predicted, color = group, group = group),
    linewidth = 1.8
  ) +
  scale_x_discrete(drop = FALSE, expand = expansion(mult = c(0, 0.02))) + 
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.15))) +
  
  scale_color_manual(values = cols) +
  scale_fill_manual(values = fills) +

  labs(x = "Year", y = "Cover (%)", color = NULL) +
  guides(
    fill = "none",
    color = guide_legend(
      override.aes = list(size = 4)
    )) +
  poster(show_legend = FALSE) +
  theme(axis.text.x = element_text(margin = margin(t = 12), size = 18))

  ggsave(
  filename = "cover.png", 
  plot = p_cover, 
  width = 9,      
  height = 4.5, 
  dpi = 300, 
  bg = "transparent" 
)

```